Clustering Constraints on the Relative Sizes of Central and Satellite Galaxies

### 2017-11-08 paper/galsize_paper.tex
Minor grammatical touchups

%---------------------------------------------------------------------------------------------------
\begin{figure*}
\centering
\includegraphics[width=\textwidth]{FIGS/sdss_small_large_sample_definitions.pdf}
\caption{
{\bf Definition of ``small" and ``large" galaxies.} For a volume-limited SDSS galaxy sample defined by $M_{\ast}>10^{10.25}M_{\odot},$ we visually demonstrate how we classify galaxies into ``small" and ``large" subsamples. As described in detail in \S\ref{subsec:sizedef}, we compute the median value $\median{\rhalf}{\mstar}$ at the stellar mass of each galaxy in the sample, and divide the sample into two based on this $\mstar-$dependent cut. The {\em left panel} compares the stellar mass functions $\phi(\mstar)$ of the two samples, confirming that our method for separating ``small" from ``large" galaxies yields subsamples with statistically indistinguishable $\phi(\mstar)$. The {\em right panel} shows histograms of $\rhalf$ for the two subsamples, which partially overlap due to the finite range of $\mstar$ in the volume-limited sample.  \textcolor{blue}{AK: x axis label in the left panel should be just $M_\star\ (M_\odot)$ without $\log_{10}$. I think this figure should be either displayed full text width like this, or made into two vertical panels and displayed in a single column. Otherwise it is too small. I wonder also whether the panel titles at the top are  needed.}
}
\label{fig:sizedefinition}
\end{figure*}
%-----------------------------------------------------------------------------------------------------

@@

As the bedrock of our modeling, we use the publicly available\footnote{\url{http://www.slac.stanford.edu/~behroozi/BPlanck\_Hlists}} catalog of {\tt Rockstar} subhalos identified at $z=0$ in the Bolshoi-Planck simulation \citep{klypin_etal11,behroozi12_rockstar,behroozi12_consistent_trees,riebe_etal13,rodriguez_puebla16_bolplanck}. As described in \S\ref{subsec:sham}, we will use traditional abundance matching to connect stellar mass $\mstar$ with subhalo peak mass $\mpeak.$ To address issues related to subhalo incompleteness \citep{guo_white13,campbell_etal17}, we supplement the Bolshoi-Planck subhalo catalog with {\em all} subhalos that were ever identified by {\tt Rockstar}, including those that no longer appear in the standard catalog as subhalos that survive to $z=0.$ We describe our treatment of these ``orphan" subhalos in the Appendix.

@@

In addition to stellar mass galaxy clustering depends on galaxy color \citep[e.g.,][]{zehavi_etal11}. Thus, we will also consider 
dependence of clustering on galaxy size while controlling for color. The results for the subsamples split in stellar mass and color are presented in \S\ref{subsec:colormorph}.

@@

\textcolor{blue}{AK: I don't think the name of the model is descriptive here, but I could not yet come up with a better one.}
In the first model we explore, we assume that stellar mass $\mstar$ controls $\rhalf,$ so that galaxy sizes are drawn from a log-normal distribution with $0.2$ dex of scatter centered at $\median{\rhalf}{\mstar}.$ To implement this model, for simplicity we tabulate $\median{\rhalf}{\mstar}$ directly from the data, rather than pursue a parametric form \citep[see, e.g.,][]{zhang_yang17}. Note that in this model the theoretical sample of model reproduces the observed $\rhalf-\mstar$ relation by construction.

@@

As noted in \S\ref{sec:intro}, \citet{kravtsov13} showed that galaxy sizes are on average proportional to the virial radius of their host halo. Motivated by these results, we  explore a model in which $\rhalf$ scales linearly with the halo virial radius:\footnote{We note that the normalization factor in \citet{kravtsov13} is 0.015, where as our normalization is 0.01. The difference derives from the choice of halo radius definition: \citet{bryan_norman98} in the present work, compared to the $200c$ definition used in \citet{kravtsov13} and the fact that the latter study used deprojected (larger) sizes for spheroidal galaxies. } 
\beq
\label{eq:fiducial_model}
\rhalf = 0.01\rvir.
\eeq
For the virial radius of halos and subhalos, we use $\rmpeak:$ the value of $\rvir$ in physical units of $\kpc$ measured at the time when halo mass reached its maximum.  The relationship between $\rmpeak$ and $\mpeak$ is given by 
\beq
\mpeak\equiv\frac{4\pi}{3}\rmpeak^{3}\Delta_{\rm vir}(\zpeak)\rho_{\rm m}(\zpeak),
\eeq
where for $\Delta_{\rm vir}(\zpeak)$ we use the fitting function to the ``virial" definition used in \citet{bryan_norman98}. For the model we refer to as the ``$\rvir-$based model", we add uncorrelated log-normal scatter of $\sigmarhalf=0.2$ dex to generate a Monto Carlo realization of the model population.

@@

Comparison of observed size dependence of galaxy clustering and model predictions in the previous section produced  two intriguing results. First,  model that assumes $\rhalf\propto\rvir$ predicts dependence of both the amplitude and shape of  galaxy correlation functions on size remarkably well, especially considering that the model has only one free parameter, normalization of the linear $\rhalf-\rvir$ relation. At the same time, small galaxies cluster more strongly relative to large, which seems at odds with naive expectations that within such model larger sizes correspond to larger halo masses, $\rvir\propto\mvir^{1/3}$, which should be clustered more strongly. Second, the strong dependence of galaxy clustering on size weakens to almost no dependence for samples of red and blue galaxies selected by color within a given stellar mass range. 
Below we propose a simple and unified explanation of all these results within the $\rvir$-based size model in \S\ref{subsec:censat_sizes}, and discuss the broader implications of our interpretation in \S\ref{subsec:broader_implications}.

@@

Recall that in the $\rvir$-based size model, galaxy size is proportional to the virial radius of host halo at the redshift at which halo had maximum mass. Thus, we can expect that in this model satellite galaxies will have sizes smaller than sizes of central galaxies hosted by halos of similar stellar mass.  Indeed, Figure \ref{fig:censatsizehist} shows the $\rhalf$ distributions of central, satellite, and ``splashback'' central galaxies hosted by halos of the same mass $\mhalo\equiv\mpeak\approx10^{12}\msun.$ A ``splashback central"  is defined as a present-day central galaxy that used to be a satellite, i.e., its main progenitor halo passed inside the virial radius of a larger halo at some point in its past history. On the other hand, we define a ``true central" as a galaxy that has never been a satellite. The figure shows that there is a clear trend of sizes decreasing from the central to splashback and to satellite galaxies.

@@

We can also understand the observed trend of size dependence of clustering with $\mstar$ (Figure \ref{fig:clustering_ratio_upshot}) in terms of the differences of sizes of central and satellites galaxies. The satellite fraction $F_{\rm sat}(\mstar)$ decreases as $\mstar$ increases \citep[e.g.,][]{guo_etal11,reddick_etal13}. As $F_{\rm sat}$ decreases, there are fewer satellites available to preferentially weight the subsample of smaller galaxies. Thus,  high $\mstar$ samples consist mostly of central galaxies and the split by size no longer corresponds to the split into satellite- and central-dominated samples. The size dependence of clustering correspondingly weakens and nearly disappears for samples with largest $\mstar$.

Likewise, the the significantly weaker dependence of clustering on size for color-selected samples (Figure \ref{fig:colorclustering}) and its reversed sign can be naturally explained within the same framework.  For example, it is well known that satellites are redder than centrals of the same stellar mass \citep[e.g.,][]{vdB_etal08}; accordingly, red galaxies cluster much strongly than  blue galaxies \citep[e.g.,][]{zehavi_etal11}. Samples of red (blue) galaxies with stellar masses of $\mstar\sim 10^{10}\ M_\odot$ are dominated by satellite (central) galaxies and split in size within such samples no longer corresponds to the split into satellite- and central-dominated samples. In this case, $\rvir$-based size model, which reproduces the observed trend, predicts that split by size does correspond to the split by halo mass $\rhalf\propto\rvir\mhalo^{1/3}$ and that that large galaxies should be somewhat more clustered than the small ones. That is they predict that size dependence of clustering  is reversed for color-selected samples and the magnitude of the dependence is significantly weaker due to weak dependence of virial radius on halo mass and weak dependence of clustering on halo mass in this regime.

While we consider such an explanation plausible and likely, a conclusive confirmation requires proper forward modeling of size and color jointly. Such an effort is beyond the scope of, but is well-motivated by the present work.

>>>>

%---------------------------------------------------------------------------------------------------
\begin{figure*}
\centering
\includegraphics[width=\textwidth]{FIGS/sdss_small_large_sample_definitions.pdf}
\caption{
{\bf Definition of ``small" and ``large" galaxies.} For a volume-limited SDSS galaxy sample defined by $M_{\ast}>10^{10.25}M_{\odot},$ we visually demonstrate how we classify galaxies into ``small" and ``large" subsamples. As described in detail in \S\ref{subsec:sizedef}, we compute the median value $\median{\rhalf}{\mstar}$ at the stellar mass of each galaxy in the sample, and divide the sample into two based on this $\mstar-$dependent cut. The {\em left panel} compares the stellar mass functions $\phi(\mstar)$ of the two samples, confirming that our method for separating ``small" from ``large" galaxies yields subsamples with statistically indistinguishable $\phi(\mstar)$. The {\em right panel} shows histograms of $\rhalf$ for the two subsamples, which partially overlap due to the finite range of $\mstar$ in the volume-limited sample.
}
\label{fig:sizedefinition}
\end{figure*}
%-----------------------------------------------------------------------------------------------------

@@

As the bedrock of our modeling, we use the publicly available\footnote{\url{http://www.slac.stanford.edu/~behroozi/BPlanck\_Hlists}} catalog of {\tt Rockstar} subhalos identified at $z=0$ in the Bolshoi-Planck simulation \citep{klypin_etal11,behroozi12_rockstar,behroozi12_consistent_trees,riebe_etal13,rodriguez_puebla16_bolplanck}. As described in \S\ref{subsec:sham}, we will use traditional abundance matching to connect stellar mass $\mstar$ with subhalo peak mass $\mpeak.$ To address issues related to subhalo incompleteness \citep{guo_white13,campbell_etal17}, we supplement the Bolshoi-Planck subhalo catalog with subhalos that have been disrupted and no longer appear in the standard catalog as $z=0$ surviving subhalos. We describe our treatment of these ``orphan" subhalos in the Appendix.

@@

In addition to stellar mass, $\wproj(\rproj)$ depends on galaxy color \citep[e.g.,][]{zehavi_etal11}. Thus, we will also consider 
dependence of clustering on galaxy size while controlling for color. The results for the subsamples split in stellar mass and color are presented in \S\ref{subsec:colormorph}.

@@

\textcolor{blue}{AK: I don't think the name of the model is descriptive here, but I could not yet come up with a better one.}
In the first model we explore, we assume that stellar mass $\mstar$ controls $\rhalf,$ so that galaxy sizes are drawn from a log-normal distribution with $0.2$ dex of scatter centered at $\median{\rhalf}{\mstar}.$ To implement this model, for simplicity we tabulate $\median{\rhalf}{\mstar}$ directly from the data, rather than pursue a parametric form \citep[see, e.g.,][]{zhang_yang17}. Note that this model reproduces the observed $\rhalf-\mstar$ relation by construction.

@@

As noted in \S\ref{sec:intro}, \citet{kravtsov13} showed that galaxy sizes are on average proportional to the virial radius of their host halo. Motivated by these results, we  explore a model in which $\rhalf$ scales linearly with the halo virial radius:\footnote{We note that the normalization factor in \citet{kravtsov13} is 0.015, where as our normalization is 0.01. The difference derives from two factors: {\em i.} the choice of halo radius definition: \citet{bryan_norman98} in the present work, compared to the $200c$ definition used in \citet{kravtsov13}; {\em ii.} the latter study used deprojected (larger) sizes for spheroidal galaxies, whereas we make no such correction here.} 
\beq
\label{eq:fiducial_model}
\rhalf = 0.01\rvir.
\eeq
For the virial radius of halos and subhalos, we use $\rmpeak:$ the value of $\rvir$ in physical units of $\kpc$ measured at the time when halo mass reached its maximum.  The relationship between $\rmpeak$ and $\mpeak$ is given by 
\beq
\mpeak\equiv\frac{4\pi}{3}\rmpeak^{3}\Delta_{\rm vir}(\zpeak)\rho_{\rm m}(\zpeak),
\eeq
where for $\Delta_{\rm vir}(\zpeak)$ we use the fitting function to the ``virial" definition used in \citet{bryan_norman98}. For the model we refer to as the ``$\rvir-$based model", we add uncorrelated log-normal scatter of $\sigmarhalf=0.2$ dex to generate a Monto Carlo realization of the model population.

@@

In the previous section, comparing model predictions to the observed size dependence of galaxy clustering produced two intriguing results. First, the model that assumes $\rhalf\propto\rvir$ predicts a dependence of both the amplitude and shape of  galaxy correlation functions on size remarkably well, especially considering that the model nominally has only one free parameter: the normalization of the linear $\rhalf-\rvir$ relation. At the same time, small galaxies cluster more strongly relative to large, which seems at odds with naive expectations that, within such a model, larger sizes correspond to larger halo masses, $\rvir\propto\mvir^{1/3}$, and so larger sizes should be clustered more strongly. Second, the strong dependence of galaxy clustering on size weakens to almost no dependence for samples of red and blue galaxies selected by color within a given stellar mass range. 
Below we propose a simple and unified explanation of all these results within the $\rvir$-based size model in \S\ref{subsec:censat_sizes}, and discuss the broader implications of our interpretation in \S\ref{subsec:broader_implications}.

@@

Recall that in the $\rvir$-based size model, galaxy size is proportional to the virial radius of host halo at the redshift at the time the halo attained its maximum mass. Thus, we can expect that in this model satellite galaxies will have sizes smaller than sizes of central galaxies hosted by halos of similar stellar mass.  Indeed, Figure \ref{fig:censatsizehist} shows the $\rhalf$ distributions of central, satellite, and ``splashback'' central galaxies hosted by halos of the same mass $\mhalo\equiv\mpeak\approx10^{12}\msun.$ A ``splashback central"  is defined as a present-day central galaxy that used to be a satellite, i.e., its main progenitor halo passed inside the virial radius of a larger halo at some point in its past history. On the other hand, we define a ``true central" as a galaxy that has never been a satellite. The figure shows that there is a clear trend of sizes decreasing from the central to splashback and to satellite galaxies.

@@

We can also understand how the observed $\rhalf-$dependent clustering scales with $\mstar$ (shown by comparing different panels in Figure \ref{fig:clustering_ratio_upshot}) in terms of the differences of sizes of central and satellites galaxies. The satellite fraction $F_{\rm sat}(\mstar)$ decreases as $\mstar$ increases \citep[e.g.,][]{guo_etal11,reddick_etal13}. As $F_{\rm sat}$ decreases, there are fewer satellites available to preferentially weight the subsample of smaller galaxies. Thus,  high $\mstar$ samples consist mostly of central galaxies, and the split by size no longer corresponds to the split into satellite- and central-dominated samples. The size dependence of clustering correspondingly weakens and nearly disappears for samples with largest $\mstar$.

Likewise, the significantly weaker dependence of clustering on size for color-selected samples (Figure \ref{fig:colorclustering}), and its reversed sign, can be naturally explained within the same framework.  For example, it is well known that satellites are redder than centrals of the same stellar mass \citep[e.g.,][]{vdB_etal08}; accordingly, red galaxies cluster much more strongly than  blue galaxies \citep[e.g.,][]{zehavi_etal11}. Samples of red (blue) galaxies with stellar masses of $\mstar\sim 10^{10}\ M_\odot$ are dominated by satellite (central) galaxies, and splitting by size within such samples no longer corresponds to the split into satellite- and central-dominated samples. In this case, the $\rvir$-based size model predicts that splitting by size {\em does} correspond to splitting by halo mass $\rhalf\propto\rvir\propto\mhalo^{1/3},$ and therefore that large galaxies should be somewhat more clustered than the small ones. That is, the $\rvir-$based model predicts that $\rhalf-$dependent clustering should be reversed for color-selected samples, and that the magnitude of the dependence should be significantly weaker due to weak dependence of virial radius on halo mass, and weak dependence of clustering on halo mass in this regime.

While we consider such an explanation plausible, and even likely, a conclusive confirmation requires proper forward modeling of size and color jointly. Such an effort is well-motivated, but beyond the scope of the present work.
